A multi-scale story of the diffusion of a new technology: the web

### 2025-01-03 paper/submission2/response.qmd

---
title: "Response to editor and reviewer comments"
format: pdf
bibliography: bibliography.bib
---

18-Dec-2024

Dear Emmanouil

Thank you for submitting your manuscript to Journal of Economic Geography.

I have now received comments from reviewers. Their comments are at the end of this email. The reviewers raise a range of concerns and have questions about the suitability of the paper for publication. The overall feeling is that the paper has probably been submitted too early and needs major work on most sections if it is to become publishable. I concur with the reviewers and see a number of significant issues that need to be addressed. This left me with a dilemma in terms of the decision. I have decided to invite you to revise and resubmit the paper so that it can be considered further for publication. However, this is a very high-risk revision, and I was bordering on rejection given the scale of the work needed and the limitations that the reviewers highlight. I am saying this to be open about the risk that after the next round the paper could be rejected unless reviewers believe you have addressed the range of issues identified. You may want to factor this into your decision about whether to revise the paper, given how much work is needed.

Having said this, let me offer some suggestions about the main areas to focus on if you decide to revise the paper. Reviewer 1 is more positive but very brief in their report. I am, therefore, putting more weight on reviewers 2 and 3 when identifying priorities for the revision.

## Editor; theory and justification 
\color{teal}
First, you will need to develop a more convincing theoretical argument and contribution. You will see that both reviewers 2 and 3 raise questions about your claims in relation to the gap in the literature and suggest you fail to build on existing work appropriately and do not convincingly identify the motivation for this analysis. In general, claims that an issue has never been addressed are hard to sustain and it is better to build on existing work and extend knowledge relating to a particular theoretical issue. The reviewers challenge you to do this using your data and suggest both the argument and structure of the paper will need to change.
\color{black}

## Editor; data
\color{teal}
Second, you need to justify your data and analysis better. Reviewer 2 fairly questions the end point of the analysis – which is not explained and given the 10+ years gap to your submission raises questions about the timeliness of the analysis. The questions about coverage of the UK website population also need to be addressed convincingly. More generally, the reviewers ask for more careful presentation of the data to make clear the significance of each figure and finding. As it stands, the paper feels rather overloaded with data, rather than acting as a focused analysis.
\color{black}

## Editor; conclusions
\color{teal}
The problems identified above result in the paper feeling imprecise, confusing and lacking a contribution. The end point is a conclusion that the reviewers do not find justified or original. The reviewers do see potential. As I noted, the feeling is a paper that is underdeveloped and submitted prematurely. There is, then, a route to publication, but it will require significant changes to convince the reviewers.
\color{black}

## Referee: 1

I have read your paper with interest and I appreciate its quality and clarity. The work is both well-executed and well-written, making it a valuable contribution to the field.

However, I would like to offer a few minor suggestions that could further enhance the readability and overall coherence of the paper:

> Simplify the Introduction: I recommend streamlining the introduction by reducing the inclusion of overly technical or highly specific details. Instead, consider expanding on the broader conceptual framework that situates the paper within its scholarly context. This approach would help orient readers and provide a more accessible foundation for understanding the study.
*Simplify the introduction*

Strengthen the Connection Between Hypotheses and Results: It may be beneficial to establish a clearer and more explicit link between the study's hypotheses and its findings. One way to achieve this could be by using section titles that are more conceptually informative, which would guide the reader through the logical progression of the argument and findings more effectively.
*change section titles?*

## Referee: 2

\color{teal}
OVERVIEW\
- focused on innovation diffusion, using the case of the web in the UK
- “this paper tests whether the three distinct spatial diffusion mechanisms shaped the diffusion of the Web in the UK at very granular spatial scales from the Web’s early stages onwards”; presumably “S-shaped pattern”/temporal models, core to periphery; distance/proximity.

Figure 2 shows the diffusion of domains per company at the scale of the UK matches the S growth model. Figure 3 shows some spatial clustering. I think this is for all years in the dataset but this could be better labeled
\color{black}

**explain, change title**

\color{teal}
Figure 4, not sure what this demonstrates
\color{black}

\color{teal}
Figure 5&6, shows early clustering and then diffusion (I think)
\color{black}

**yes**

\color{teal}
DATA\
Focusing just on .co.uk website is OK but needs to be better justified given that it only represented about 50 percent of domains in the UK at that time. See my back of the envelop calculations: As per Zook (2001, Table 1) .uk represented 66% of all domains in the UK, .co.uk is a subset of that, so this is perhaps 50% of all domains registered in the UK at that time. If 1/5 of .co.uk domains had findable postal codes (p. 5), this study is based on 10% of the domains in the UK. (perhaps this is too low, aggregating from webpages to websites brought the number higher, not entirely clear to me)
\color{black}

Thanks for going the extra mile to check the data used here! I agree with the 
above calculation up until the findable postcode point as 1/5 of the archived 
*webpages* -- not *websites*, i.e. domains -- included a postcode in the web text. 
Think of the website of a small company with one trading address. I think it is 
fair to expect that website to have multiple webpages to present the different
products and services and then one webpage to showcase the address -- the 
'Contact us' page. So, the 1/5 share of *webpages* containing a postcode should
translate to a much higher share of *websites* -- i.e. domains -- containing 
a postcode. Previous estimates indicated that around 70% of all websites contain 
place reference [@hill2009georeferencing] and this reference has no been added 
in the paper.

So, based on the above 35% of all domains with a registered address in the UK
were represented in the data. Most importantly though the paper does not model 
the adoption of the Web from the general population but instead from firms. So, 
although the data only represents 35% of the population of websites with a 
registered address in the UK, it represents almost all the *relevant* archived
websites -- that is websites using the commercial .co.uk SLD. As stated in the 
paper, this has been the primary choice for commercial websites in the UK.

The third paragraph in the Data and Methods section includes the following text:

> Instead of using the whole .uk ccTLD, this paper focuses on its
commercial subset, the .co.uk second level domain (SLD). This choice
is aligned with the topic of the paper -- the diffusion of the Web for 
commercial functions and decreases the heterogeneity of the web data as such 
commercial websites have specific aims: they are used to diffuse information, 
support online transactions and share opinions [@THELWALL2000441; 
@blazquez2018big].

<!-- Active/inactive, don;t suffer from multiple registrations -->

\color{teal}
overall the time period 1996-2012 is a good time to study the diffusion of the commercial internet. But why stop at 2012? Is it when the Library stopped its data series?
\color{black}

Yes, the British Library had obtained a copy of all of the .uk contents of the 
Internet Archive for that period. From a data perspective, it wouldn't be correct
to extend this dataset with data from other open archives such as the Common 
Crawl as they use different crawlers and there will be inconsistencies. Most
importantly though, this period covers the diffusion of the web from the very
early stages until its maturity. This is also illustrated by the figure 
depicting the cumulative adoption and the flattening right end of the line.

There is a specific paragraph justifying how the time period of the data fits
the aim of the paper and another line is now included in the second paragraph 
of the introductions ("from the very early days until its maturity (1996-2012)").

\color{teal}
Framing
“With the exception of Sinai and Waldfogel (2004), who focused, among other things, on web content and demonstrated that more such content is available in larger markets, this strand of literature has been mostly concerned with the spatiality of the Internet’s hard and tangible infrastructure and shied away from exploring the spatial dimensions of one of the most distinct software layers of the Internet infrastructure, the Web.” This statement is simply not true, the literature reviews misses the entirety of web content diffusion studies: Zook, 2000, Malecki, 2002. Townsend, 2000, Gorman, 2002, Feldman, 2002, lots more
\color{black}

Point taken. Despite being aware of these studies, I omitted citing them as I 
was primarily searching for studies testing the diffusion mechanisms over longer
time periods. This error has now been corrected and the above paragraph has been 
replaced with the following one, which emphasises the long time series motivation. 
Regarding the actual references, @zook2000web, @zook2000web and @moss1997tracking 
used relevant, domain name registration data, but the rest (e.g. @malecki2002economic 
and @moss2000internet) focused on hard infrastructure. @gorman2002web examined 
the head and branch locations of the top 40 e-business integration firms in the 
USA, which although interesting is not directly relevant to this paper. I also
went through of Feldman's work around 2002, but I did not find papers
adopting a similar empirical perspective.

> Another line of motivation has to do with the very nature of the Web as a 
digital technology and how it evolved over time. The digital geographies field 
has debunked premature claims about the Internet's placeless-ness by mapping 
various of its facets and technological layers [@tranos2013death], but the lack 
of relevant long time-series data prevented researchers from observing digital
phenomena overtime. Most of these studies focused on its hard infrastructure 
[e.g. @malecki2002economic; @moss2000internet], on Internet subscribers 
[e.g. @blank2018local], or on social media users [e.g. @crampton2013beyond].
Exceptions include early work from @zook2000web, @zook2000web and 
@moss1997tracking, which utilised domain name registration data and use it to
map and illustrate the spatial footprint of the web content. Pivotal as these 
contributions have been in understanding the economic geographies of the 
Internet, they were based on cross-sectional, often spatially detailed, mapping 
and on shallow time comparisons (e.g. 1993-1998) and did not observe the 
evolution and diffusion of the Web until its maturity. As we know from the 
relevant literature, different technologies diffuse at different rates and 
there is no theory that can explain these differences 
[@leibowicz2016representing]. Hence, mapping and modelling the diffusion of the
Web among firms can offer useful insights.

\color{teal}
“for a digital technology as weightless as the Web” – this runs a bit counter to lots of work by digital geographers who have shown again and again how the digital is not weightless or placeless
\color{black}

\color{teal}
Conclusion
p. 18 “This is the first time that the importance of granular spatial mechanisms for the diffusion of an intangible, digital technology is exposed.” Honestly, I would not expect the diffusion of the web to be any different than any other kind of innovation diffusion. The spatiality of digital life, and its connection to material spaces is a constant refrain within digital geography. So I’m not convinced it makes much sense to emphasize this, if a granular spatial analysis of innovation diffusion is already novel.
\color{black}

**for the last two comments, I need to soften the weightless argument throughout.**
I have added a long time series argument and also said that there is heterogeneity
[@leibowicz2016representing] among diffusion rates of different technologies.

As per the 'framing' point above, the digital motivation para in the 
introduction has been amended acordilngly.

## Referee: 3

\color{teal}
This paper addresses a gap in Economic Geography by examining the spatial diffusion of digital technologies, specifically the adoption of commercial websites, from 1996 to 2012. Recognizing the potential of digital technologies to shape local development trajectories through increased productivity and innovation, the study employs novel data sources and machine learning techniques to analyze diffusion patterns at highly granular geographical scales. The authors explore the roles of various spatial mechanisms driving this process, offering insights into how digital technology adoption unfolds across space and time.
The paper has the potential to make an interesting contribution to the literature as it combines a fascinating topic with novel data and methodologies. However, in its current form, it is not ready for publication, as it never becomes clear what it actually delivers in terms of new knowledge and what knowledge gap it seeks to fill. Despite some analytical elements, it is rather descriptive in nature, which doesn’t fit well with this journal.

Most importantly, I suggest developing some clear hypotheses to structure and motivate the paper better. In light of this, I recommend major revisions.
\color{black}

This is very helpful. The first two sections have been restructured around two 
hypotheses. I believe the structure and the flow of the paper is much clearer
now.

\color{teal}
\noindent 1. Abstract:\
\noindent 1. The abstract needs significant improvement. It needs to adequately report the main findings, relevance, or implications of the study. Additionally, the assertion that geographers have "shied away" from analyzing the diffusion of digital technologies is misleading; the issue was primarily due to the lack of available data.
\color{black}

**fix the abstract**

"shied away" has been replaced in the opening paragraph.

\color{teal}
\noindent 2. Introduction\
\noindent 2. General Diffusion Studies: The statement that there are few contemporary general diffusion studies in Economic Geography is not accurate, although it may apply specifically to digital technologies. Consider the following recent examples:
o Peris, A., Meijers, E., & van Ham, M. (2021). Information diffusion between Dutch cities: Revisiting Zipf and Pred using a computational social science approach. Computers, Environment and Urban Systems, 85(November), 101565. https://doi.org/10.1016/j.compenvurbsys.2020.101565
o Bednarz, M., & Broekel, T. (2020). Pulled or pushed? The spatial diffusion of wind energy between local demand and supply. Industrial and Corporate Change, 29(4), 893–916. https://doi.org/10.1093/icc/dtaa012
o Feldman, M. P., Kogler, D. F., & Rigby, D. L. (2015). rKnowledge: The Spatial Diffusion and Adoption of rDNA Methods. Regional Studies, 49(5), 798–817. https://doi.org/10.1080/00343404.2014.980799
\color{black}

@PERIS2021101565 is indeed an interesting paper, but it is not directly relevant 
as it focuses on information diffusion rather than diffusion of new 
technologies. Information diffuses through social and, consequently, urban 
networks, but it is not adopted like new technologies. @bednarz2020pulled was 
already cited and @feldman2015rknowledge has now been added. These references 
are included in the second section.

Regarding the statement about economic geography studies on diffusion of 
new technologies and innovation, I do believe that this is a fair statement. 
The references listed below are the ones reviewed in the second section and none 
of them are published in one of the flagship economic geography journals 
(*Journal of Economic Geography* and *Economic Geography*). There are two papers
published in *The Professional Geographer*, which is a broad-scope geography 
journal and then only one paper in *Regional Studies*, which is the closest to
economic geography. Importantly though, the *Regional Studies* paper 
[@feldman2015rknowledge] paper is focused on inventions as observed by patent 
data contrary to the broader 'adoption of a new technology' approach of this 
paper. All other papers listed below are published in business and economics 
journals.

1. @beardsell1999spatial: European Economic Review
2. @bednarz2020pulled: Industrial and Corporate Change
3. @haller2011determinants: Applied Economics
4. @perkins2011internet: The Professional Geographer
5. @feldman2015rknowledge: Regional Studies
6. @ding2010modeling: The Professional Geographer
7. @bakher2013diffusion: Journal of Marketing Analytics
8. @PAPAGIANNIDIS2015308: Technological Forecasting and Social Change
9. @lengyel2020role: Scientific reports

Also, the lack of spatial granularity argument has been highlighted in various 
instances throughout the paper as the main line of justification.

**ADD SPATIAL GRANULARITY IN CONCLUSIONS**

\color{teal}
\noindent 3. Novelty: The study needs to be more explicit about its novel contributions to understanding spatial diffusion mechanisms. Currently, the discussion is too general.
\color{black}

**granularity**

**HAVE A THINK, intro has been restructured**

\color{teal}
\noindent 4. Motivation: The motivation for the study is unclear. The text raises several issues, but it is unclear which are specific to digital technologies and which pertain to technological diffusion more broadly.
\color{black}

**HAVE A THINK, intro has been restructured**

\color{teal}
\noindent 5. Structure and Focus: The discussion on motivations should be streamlined and clarified. For example, after specifying that the study focuses on commercial websites rather than the web, subsequent references revert to the general term “web.” Additionally, the discussion confuses diffusion patterns, firm perspectives, and early adopter advantages. Consider restructuring this section to clearly differentiate between literature gaps (motivations) and expected outcomes (mechanisms).
\color{black}

The revised paper consistently illustrates that the focus is on commercial 
websites.

In the relevant paragrapg in the Data and Methods section, the following line
has been added: "Henceforward any reference to websites will refer to commercial 
-- .co.uk -- websites".

**fixed for intro, lit rev  and data -- check the rest of the paper.**

The introduction and literature review sections have been restructured, so 
the firm perspectives and early adopter advantages discussions have been moved.

\color{teal}
\noindent 6. Data Consistency: The terminology for the data should be consistent. At one point, the text refers to the “totality of the web,” while elsewhere, it specifies “commercial websites.” Clarify and maintain consistency throughout.
\color{black}

The terminology is now consistent across the paper. The "totality" point 
referred to different web technologies and not to the totality of the web. It 
has now been replaced with "an overarching technology".

\color{teal}
\noindent 7. Recommended Structure: To align with the common structure in Journal of Economic Geography (JoEG), consider splitting the introduction into a shorter introduction and a dedicated theory section merged with the literature review.
\color{black}

The introduction is now shorter and the literature review section has been 
re-structured and re-named as "The mechanisms that shape the diffusion of a new 
technology".

\color{teal}
\noindent 3. Literature Review / Theory\
\noindent 1. Synthesis and Motivation: The literature review should be transformed into a theory section. Rather than presenting a historical review, synthesize the arguments to clearly motivate and inform your hypotheses.
\color{black}

Done. Restructured: theory, gaps, hypotheses.

\color{teal}
\noindent 2. Knowledge Gaps: The current review does not clearly identify the knowledge gaps that this study aims to address. Clarify whether the goal is to confirm known processes with new data and methods or to explore unique diffusion patterns related to websites.
\color{black}

Known processes at very detailed scale: explanatory power of these mechanisms at granular, local
scales

\color{teal}
\noindent 3. Lack of Hypotheses: To clarify what the paper delivers and what knowledge gaps it fills, I strongly recommend developing a number of hypotheses that will be tested.
\color{black}

Added.

\color{teal}
\noindent 4. Fit to title: The term “multi-scalar” appears in the title but is never really discussed in the paper.
\color{black}

This was discussed from the data section onwards, but indeed not before that. 
Section 2 includes a new paragraph describing the second hypothesis (H~2~), 
which explicitely discusses the value of the multiscale approach: 
"Understanding the diffusion of a new technology in multiple, detailed spatial
scales is of interest to economic geography as ..."

\color{teal}
\noindent 4. Data & Methods\
\noindent 1. Order of Presentation: Introduce the data first, followed by the geographical units used in the study.
\color{black}

The order of presentation has been amended accordingly and the two scales of 
analyses are presented after the data.

\color{teal}
\noindent 2. Postcode Identification: The approach of identifying postcodes on websites is innovative, but potential errors need to be discussed. Specifically:
- How do you ensure that a postcode on a website accurately represents the 
company’s location or its subsidiary?
- How do you handle numbers that resemble postcodes but are not actual postcodes?
- What about references to events or other organizations unrelated to the website’s owner?
Include a brief discussion of these potential issues.
\color{black}

A number of different strategies have been developed to address the potential 
issues described above.

\noindent 1. To prevent false positives, the resulted postcodes
were matched against official postcode lookup files [@ons]. This was not 
explicitly mentioned in the first submission on the paper and the previous 
sentence has now been added in the relevant paragraph:

### 2025-01-03 paper/submission1/anonymised.qmd
ssh problem

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
  This paper maps the participation in the digital economy and its evolution in the UK over space and time. Most of the existing economic geography literature which dealt with the spatiality of the internet employed supply-side measures, such as infrastructural capacity, in order to understand the geography of the digital economy and its potential spatial economic effects. Useful as these approaches might have been, they cannot capture the micro-processes and the characteristics of the individual online behaviour. Using large volumes of archived and geolocated web content, this paper models the diffusion of web technologies over space and time in the UK. The data and geolocation strategy allow to capture these processes at a very granular spatial scale. The modelling approach, which is based on simple spatial analytical methods and on the estimation of diffusion curves at various scales, enables to depict the role of geography and other cognitive factors which drove the diffusion of web technologies. Although the focus is on a recent historical period -- 1996-2012 -- the results of the analysis depict diffusion mechanisms which can be very useful in understanding the evolutionary patterns of the adoption of other newer technologies.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

<!-- geographers lack of interest for diffusion + motivation -->
Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of @hagerstrand1968innovation,
which demonstrated the importance of communications as a mechanism for the 
diffusion of innovations, being the most impactful. It was followed by various 
empirical studies -- see for instance @ormrod1990 and @iso2005. 
However, the torch of exploring and modelling such diffusion processes had been
passed to other disciplines such as economics, business studies and
sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the recent lack of
geographical studies exploring the diffusion of new and, more specifically for 
this paper, digital technologies across both space and time can be attributed to 
the scarcity of relevant and granular enough data, a problem also highlighted 
by @iso2005 and @kemeny2011international. As @zook2022mapping pointed, digital 
activities are hardly ever captured in official data.

<!-- contribution -->
This paper offers such a contribution: a geographical study demonstrating
how a new technology that is the Web diffused over space and time in the
UK at a high level of spatial granularity during the 1996-2012 period.
It does so by employing a novel source of big data and machine 
learning algorithms to model the *active* engagement with Web during that 
period and tests how different spatial and temporal mechanisms shaped its 
diffusion. This paper also exemplifies how the combination of data sources 
which escape the traditional social science domain and adequate research 
methods can offer new lenses to geographical research regarding the 
understanding of technological diffusion.

<!-- motivation: stakeholders -->
The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
about the diffusion of related *future* technologies. As per
@leibowicz2016representing, historical studies agree that technologies diffuse 
differently in terms of times, rates, and geographies and can be driven
by related policies [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers, network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there is a scarcity 
of empirical studies analysing the diffusion of new digital technologies 
at a detailed geographical level.

<!-- motivation: regional development -->
This gap becomes even more prominent if we consider the importance of 
technology and its adoption in economic geography and macroeconomic growth 
theories. Simply put, differences in technological adoption and sophistication 
may lead to uneven development as new technologies can increase productivity 
and allow firms to introduce new products [@solow1957technical; 
@aghion1990model; @kemeny2011international]. Linked to the subject of this 
paper, the Web -- and the active engagement with this technology that is 
creating instead of just browsing a website -- can be considered as the 
underpinning technology for the transformative process known as the
*digital service economy* [@capello2024nexus].
After all, diffusion together with invention and innovation are considered 
the pillars of technological change [@das2022diffusion].
Findings from previous empirical work illustrated that early engagement with 
digital technologies such as the Web led to longstanding positive productivity 
gains for these regions that hosted early adopters [@tranosuk].

<!-- motivation: intangible/digital -->
Another line of motivation has to do with the intangible nature of the Web 
and the debate about the footloose nature of digital technologies, which
contrary to several earlier general purpose technologies are light in weight 
[@coyle1999weightless]. Following the seminal work of @gaspar1998information, 
which demonstrated that technological improvements in telecommunications lead 
to increased demand for face-to-face communications and, consequently, 
enhance the importance of cities, empirical studies from economic geography 
and urban economics explored whether digital technologies substitute or 
complement agglomeration externalities and spatial structure 
[@kolko_death_2000; @ioannides2008effect; @tranos2021ubiquitous].
With the exception of @sinai2004geography, who focused, among other things, 
on web content and demonstrated that more such content is available in larger
markets, this strand of literature has been mostly concerned with the
spatiality of the internet's hard and tangible infrastructure
and shied away from exploring the spatial dimensions of one of the 
most distinct software layers of the internet infrastructure, the Web.

@@

<!-- early engagement with the web -->
This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.[^1] Also, only
9 per cent of UK's households had access to the internet in 1998
[@ons2018], the web included mostly static webpages, there were no
social media and web browsing involved exclusively a desktop PCs
as there were no smartphones [@tranosuk]. Hence, it is fair to say that
the study period captures the very early stages of the diffusion of a
new technology that is the Web as well until its maturity. 
The former is a key point in the lifecycle of a new technology. 
Firstly, during this period new technologies are expensive, crude and 
imperfect [@rosenberg1994exploring; @wilson201281].
A simple comparison between Web 1.0 and Web 3.0 applications clearly
illustrates this [@tranos2020social]: while a static website facilitates
one-way information dissemination, a platform like `github` enables 
cooperation between users and the creation of new information, meaning, 
and even knowledge [@faraj2016special; @barassi2012does]. 
During this period the performance of a new technology is the main 
attraction and not the cost to access and use it [@wilson2011lessons]. 
There is a broader theoretical discussion in the literature about 
the early adoption motives. As summarised by @perkins2005international, 
on one hand, epidemic models highlight the role of interpersonal 
contacts as a way for new technologies to diffuse. On the other hand, 
economic models underline the importance of heterogeneity. Different firms have
different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories [@neffke2011regions; @simmie2014new]. This
argument is even more relevant when the focus is on digital technologies because 
of the commonly found lag between investment and economic returns as reflected in 
the Solow paradox [@acemoglu2014return; @brynjolfsson2018artificial].

@@

<!-- data -->
The data used here depict the active engagement with the Web as they contain 
geolocated and time-stamped counts of commercial websites from the UK. Instead 
of adopting metrics reflecting the passive engagement with the internet such as 
internet adoption and speeds [e.g. @blank2018local; @destefano2022fuzzy], this 
paper observes yearly counts of geolocated websites, which reflect the creation 
and maintenance of websites. It needs to be highlighted here that although 
there are various types of underpinning web technologies -- namely markup, 
stylesheet, client- and server-side coding languages as well as different
web browsers and other related programming language -- the focus here
is on the diffusion of the Web in its totality. For a more detailed and 
cross-country technological diffusion study, see @PAPAGIANNIDIS2015308.

>>>>

---
title: "A multi-scale story of the diffusion of a new technology: the web"
abstract: |
    Despite the broad consensus that differences in technological adoption and sophistication can lead to uneven trajectories of local development as new technologies can increase productivity and allow firms to introduce new products and services, economic geographers have shied away from exploring the diffusion of digital technologies across space and time. This paper fills this gap by exposing the roles of different spatial mechanisms in the diffusion of an intangible, digital technology at very granular geographical scales. It does so by employing a novel source of data and machine learning algorithms to model the active engagement with Web during 1996-2012.
keywords: 
  - diffusion
  - technologies
  - Web
  - digital
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

<!-- geographers lack of interest for diffusion + motivation -->
Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of 
@hagerstrand1968innovation, which demonstrated the importance of communications 
as a mechanism for the diffusion of innovations, being the most impactful. 
It was followed by various empirical studies -- see for instance @ormrod1990 
and @iso2005. However, the torch of exploring and modelling such diffusion 
processes had been passed to other disciplines such as economics, business 
studies and sociology well before the 'cultural turn' of economic geography
[@perkins2005international]. A potential explanation of the lack of
geographical studies exploring the diffusion of new and, more specifically for 
this paper, digital technologies across both space and time can be attributed to 
the scarcity of relevant and granular enough data, a problem also highlighted 
by @iso2005 and @kemeny2011international. As @zook2022mapping pointed, digital 
activities are hardly ever captured in official data.

<!-- contribution -->
This paper offers such a contribution: an economic geography study revealing
the importance of spatial mechanisms for the diffusion of an intangible, 
digital technology that is the Web in the UK at a high level of spatial 
granularity during the 1996-2012 period.
<!-- how a new technology that is the Web diffused over space and time  -->
It does so by employing a novel source of big data and machine 
learning algorithms to model the *active* engagement with Web during that 
period and tests how different spatial and temporal mechanisms shaped its 
diffusion. This paper also exemplifies how the combination of data sources 
and state-of-the-art research methods which escape the traditional social 
science domain can offer new lenses to geographical research regarding the 
understanding of technological diffusion.

<!-- motivation: stakeholders -->
The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffuse over space and time and use this knowledge to make predictions
about the diffusion of related future technologies. As per
@leibowicz2016representing, historical studies agree that technologies diffuse 
differently in terms of times, rates, and geographies and can be driven
by related policies [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers, network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there is a scarcity 
of studies analysing the diffusion of new digital technologies at a detailed
geographical level because of the lack of detailed enough data and adequate 
research methods.

<!-- motivation: regional development -->
This gap becomes even more prominent if we consider the importance of 
technology and its adoption in economic geography and macroeconomic growth 
theories. Simply put, differences in technological adoption and sophistication 
may lead to uneven development as new technologies can increase productivity 
and allow firms to introduce new products [@solow1957technical; 
@aghion1990model; @kemeny2011international]. Linked to the subject of this 
paper, the Web -- and the active engagement with this technology that is 
creating and maintaining instead of just browsing a website -- can be 
considered as the underpinning technology for the transformative process 
identified by @capello2024nexus as the *digital service economy*. Importantly, 
the focus here is not the Web in general, but instead commercial websites.
After all, diffusion together with invention and innovation are considered the 
pillars of technological change [@das2022diffusion]. Findings from previous 
studies illustrated that early engagement with digital technologies such as 
the Web led to longstanding positive productivity gains for these regions that 
hosted early adopters [@tranosuk]. Hence, obtaining a robust understanding of 
the spatial diffusion mechanisms is crucial in mapping future development 
trajectories.

<!-- motivation: intangible/digital -->
Another line of motivation has to do with the intangible nature of the Web 
and the debate about the footloose nature of digital technologies, which
contrary to several earlier general purpose technologies are 'light in weight' 
[@coyle1999weightless]. Following the seminal work of @gaspar1998information, 
which demonstrated that technological improvements in telecommunications lead 
to increased demand for face-to-face communications and, consequently, 
enhance the importance of cities, empirical studies from economic geography 
and urban economics explored whether digital technologies substitute or 
complement agglomeration externalities and spatial structure 
[@kolko_death_2000; @ioannides2008effect; @tranos2021ubiquitous].
With the exception of @sinai2004geography, who focused, among other things, 
on web content and demonstrated that more such content is available in larger
markets, this strand of literature has been mostly concerned with the
spatiality of the Internet's hard and tangible infrastructure
and shied away from exploring the spatial dimensions of one of the 
most distinct software layers of the Internet infrastructure, the Web.

@@

<!-- early engagement with the web -->
This paper focuses on the diffusion of the Web as new technology during
the 1996-2012 period. This was an exciting period for digital
technologies as it corresponds with the commercialisation of the
Internet and, consequently, its almost universal adoption. The reader is
reminded that it was only in 1994 when Netscape Navigator was
introduced, a year before Microsoft's Internet Explorer.[^1] Also, only
9 per cent of UK's households had access to the Internet in 1998
[@ons2018], the web included mostly static webpages, there were no
social media and web browsing involved exclusively a desktop PCs
as there were no smartphones [@tranosuk]. Hence, it is fair to say that
the study period captures the very early stages of the diffusion of a
new technology that is the Web until its maturity. The former is a key 
point in the lifecycle of a new technology. Firstly, during this period new 
technologies are expensive, crude and imperfect [@rosenberg1994exploring; 
@wilson201281]. A simple comparison between Web 1.0 and Web 3.0 applications 
clearly illustrates this [@tranos2020social]: while a static website 
facilitates one-way information dissemination, a platform like `github` 
enables cooperation between users and the creation of new information, 
meaning, and even knowledge [@faraj2016special; @barassi2012does]. 
During this period the performance of a new technology is the main 
attraction and not the cost to access and use it [@wilson2011lessons]. 
There is a broader theoretical discussion in the literature about 
the early adoption motives. As summarised by @perkins2005international, 
on one hand, epidemic models highlight the role of interpersonal 
contacts as a way for new technologies to diffuse. On the other hand, 
economic models underline the importance of heterogeneity. Different firms 
have different structures and business plans, which define the potential
economic returns of the adoption of a new technology and, therefore, the
choice to adopt a new technology becomes an individual option. From a
broader and evolutionary perspective, initial conditions are essential
for the creation and evolution of path-dependent technological
development trajectories [@neffke2011regions; @simmie2014new]. This
argument is even more relevant when the focus is on digital technologies 
because of the commonly found lag between investment and economic returns as 
reflected in the Solow paradox [@acemoglu2014return; 
@brynjolfsson2018artificial].

@@

<!-- data -->
The data used here depict the active engagement with the Web as they contain 
geolocated and time-stamped counts of commercial websites from the UK. Instead 
of adopting metrics reflecting the passive engagement with the Internet such as 
Internet adoption and speeds [e.g. @blank2018local; @destefano2022fuzzy], this 
paper observes yearly counts of geolocated websites, which reflect the creation 
and maintenance of websites. It needs to be highlighted here that although 
there are various types of underpinning web technologies -- namely markup, 
stylesheet, client- and server-side coding languages as well as different
web browsers and other related programming language -- the focus here
is on the diffusion of the Web in its totality. For a more detailed and 
cross-country technological diffusion study, see @PAPAGIANNIDIS2015308.

<!-- description of the methods and results -->
To analyse these data, a diverse set of methods are employed. Firstly, 
cumulative adoption curves are estimated for the UK as a whole and for the 
Local Authority Districts (LAD), illustrating, for the first time, such 
S-shaped diffusion patterns for local areas. This system level analysis also 
reveals insights about the volatility of the adoption of the Web. Then, 
exploratory spatial data analysis offers some first indications of the 
importance of the different spatial diffusion mechanisms in shaping the 
diffusion of the Web in the UK. Importantly, the analysis adopts two distinct 
geographical scales: the LAD and the most detailed scale of UK census areas, 
the Output Areas (OA). Lastly, a novel methodological framework based on 
Machine Learning (ML) algorithms is developed to test the 
role of such mechanisms at these two scales. The results reveal, for the first 
time, the importance of different spatial diffusion mechanisms in the diffusion 
of an intangible, digital technology at very granular geographical scales.

<!-- contents -->
The structure of the paper goes as follows. @sec-litreview reviews the key
concepts in the diffusion literature as well as relevant empirical studies. 
Then, @sec-datamethods describes the data and methods used in this paper. 
The results of the different stages of the analysis are presented in 
@sec-results. The discussion of the results and the conclusions of the paper
are offered in @sec-conclusions.
